---
title: "LinUCB bandits"
author: "Romain Ferrali"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LinUCB bandits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: lib.bib
---

The introductory vignette gives a short introduction to bandit algorithms, and their implementation in the `banditr` package. This vignette details the specifics of how `banditr` handles Linear Upper Confidence Bounds (LinUCB, [@Li2010a]) bandit algorithms. 

Let $\hat{y}_{it}$ be the predicted outcome for treatment arm $i$. At each time period $t$, the LinUCB algorithm selects the experimental arm $i$ that maximizes $p(i,t) = \hat{y}_i^T \beta_t + \alpha U_t(i)$, where $U_t(i) = \sqrt{z_i^T\left(X_t^T X_t + \lambda I_K \right)^{-1} z_i}$. The parameter $\alpha \geq 0$ is a tuning parameter that arbitrates between exploitation (selecting treatment arms with a high predicted reward $\hat{y}_i$), and exploration (selecting treatment arms with high uncertainty $U_t(i)$). The parameter $\lambda > 0$ introduces ridge regularization. 

For modelling, the package currently supports linear, and logistic regressions with ridge regularization, and allows for a first stage of variable selection using the LASSO, as well as the automatic selection of tuning parameters using $K-fold$ cross-validation. All these functionalities are implemented using the `glmnet` package, wrapped into the new `banditGlmnet` class, that supports the formula interface. 

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
